---
title: "Lecture 19 Treatment Effect Methods and Review"
author: "Nick Huntington-Klein"
date: "`r Sys.Date()`"
output:   
  revealjs::revealjs_presentation:
    theme: solarized
    transition: slide
    self_contained: true
    smart: true
    fig_caption: true
    reveal_options:
      slideNumber: true
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
library(tidyverse)
library(dagitty)
library(ggdag)
library(gganimate)
library(ggthemes)
library(ggpubr)
library(modelsummary)
library(fixest)
library(Cairo)
theme_set(theme_gray(base_size = 15))
```

## Some Pointers

- Last time we talked about heterogeneous treatment effects and how our methods produce different averages of those effects
- But we don't need to be limited to that!
- There are plenty of methods - many of them new - that let us estimate a *distribution* of treatment effects
- We won't be going super far into detail with them, but I'll mostly just be letting you know they exist and some pointers for looking further
- I'll favor pointers to packages over papers, but if you look in the help files you'll generally find paper citations

## Sorted Effects

- The *sorted effects* method uses covariates to look at variation in the treatment effect, and produces a distribution of treatment effects
- It also lets you see *who* is at each part of the distribution

```{r, echo = TRUE, results='hide'}
library(SortedEffects)
# Data on being denied for a mortgage.
data(mortgage)
# Save the formula to reuse later
fm <- deny ~ black + p_irat + hse_inc + ccred + mcred + pubrec + ltv_med + 
           ltv_high + denpmi + selfemp + single + hischl
# spe() for "Sorted Partial Effects"
m <- spe(fm, data = mortgage,
         var = 'black', # black is the treatment variable
         method = 'logit',
         us = c(2:98)/100, # Get the distribution from the 2nd to 98th percentile
         b = 500, bc = TRUE) # Use bootstrapped SEs and bias-correction 
```

## Sorted Effects

- See the distribution in the effect of black on being denied

```{r, echo = FALSE}
plot(x = m)
```

## Sorted Effects

- Who is most and least affected by the "treatment" of being black?

```{r, echo = TRUE}
t <- c("deny", "p_irat", "black", "hse_inc", "ccred", "mcred", "pubrec",
       "denpmi", "selfemp", "single", "hischl", "ltv_med", "ltv_high")
classify <- ca(fm, t = t, data = mortgage, var = 'black', method = 'logit',
               cl = 'both', # Get WHO the most and least are, not how different they are
               b = 500, bc = TRUE)
results <- summary(classify) %>%
  as_tibble() %>%
  mutate(Group = row.names(summary(classify)),
         Ratio = Most/Least) %>%
  select(Group, Most, Least, Ratio) %>%
  arrange(Ratio)
```

## Sorted Effects

- Those who were denied for insurance (`denpmi`) had smallest effects of `black`, those who were `single` had the biggest

```{r, echo = FALSE}
knitr::kable(results)
```

## Bayesian Hierarchical Modeling

- A very old method! But it works. An extension of random effects
- Instead of just letting the *constant* vary, let *any* coefficient vary, and give each its own function to vary over controls! Those controls can let the effect vary

$$ Y = \beta_0 + \beta_1X + \varepsilon $$
$$ \beta_0 = \gamma_{00} + \nu_{00} $$
$$ \beta_1 = \gamma_{10} + \gamma_{11}W + \nu_{01} $$

- Terminology difference: "fixed effects" means "coefficients that don't vary"

## Bayesian Hierarchical Modeling

```{r, echo = TRUE}
library(lme4)
# The whole thing would be super slow so for now let's just do a few effects
m <- lmer(deny ~ p_irat + hse_inc + ccred + mcred + pubrec + ltv_med + 
           ltv_high + denpmi + selfemp + single + hischl +
       (single + hischl | black),
     data = mortgage)
```

## Bayesian Hierarchical Modeling (cut off)

```{r, echo = FALSE}
summary(m)
```

## Machine Learning

- The biggest contributions of machine learning to causal inference thus far have been in heterogeneous treatment effects
- (there are other things too, like matrix completion, which is way too complex to get into here)
- Allowing a zillion different things to vary is easy in machine learning!
- Note: machine learning tends to use "training" and "holdout" data. So estimate the model using a training subset, and then estimate your treatment effect distribution by sending your holdout data through that model

## LASSO and interactions

- LASSO is a "regularized regression" that does regression but doesn't JUST minimize sum of squared errors, it also has a second goal of shrinking coefficients
- There are several forms of regularized regression. LASSO tends to set coefficeints to 0, i.e. chuck them out
- So... just interact treatment with everything and see what interactions are worth keeping!

## LASSO and interactions

- This is commonly applied in instrumental variables settings, where these interactions can improve first-stage power and ease the weak-instrument problem
- But can be applied elsewhere too to find variation in a treatment effect
- I won't do a walkthrough here, but the package **glmnet** is typically used to estimate LASSO. See [this walkthrough](https://www.statology.org/lasso-regression-in-r/).

## Causal Forest

- A *random forest* is a prediction method. Take your data, go through every covariate you have, and every \emph{value} of each covariate, and split the data based on the split that minimizes sum of squared error
- Then do that again for each of the splits, and again and again. Each time only use a random subset of the variables
- Stop once the splits get too small
- *Causal forest* does the exact same thing except instead of minimizing the SSE, it *maximizes* the difference in causal effect between the splits
- i.e. it hunts for causal effect differences! You end up with an effect prediction for each individual

## Causal Forest

- **grf** does causal forest, and even has an IV version if you need an IV to identify

```{r, echo = TRUE}
library(grf)
mortgage <- mortgage %>% mutate(holdout = runif(n()) > .5)
holdout <- mortgage %>% filter(holdout)
training <- mortgage %>% filter(!holdout)
W = training %>% pull(black) %>% as.matrix()
X = training %>% select(p_irat, hse_inc, ccred, mcred, pubrec,
       denpmi, selfemp, single, hischl, ltv_med, ltv_high) %>% as.matrix()
Y = training %>% pull(deny) %>% as.matrix()
m <- causal_forest(X, Y, W, tune.parameters = 'alpha')
```

## Causal Forest

```{r, echo = TRUE}
X.holdout <- holdout %>% select(p_irat, hse_inc, ccred, mcred, pubrec,
       denpmi, selfemp, single, hischl, ltv_med, ltv_high) %>% as.matrix()
indiv_effects <- predict(m, X.holdout)
holdout <- holdout %>% mutate(effect = indiv_effects$predictions)
```

## Causal Forest

```{r, echo = FALSE}
ggplot(as_tibble(holdout), aes(x = effect)) + geom_density() + 
  theme_pubr() + 
  labs(x = 'Individual Effect of Black on Denial Probability',
       y = 'Density')
```

## Causal Forest

- Who is affected? Let's do a similar test to what **SortedEffects** did (although we could look at it plenty of other ways)

```{r}
holdout %>%
  mutate(Range = case_when(
    effect <= quantile(effect, .05) ~ 'Bottom',
    effect >= quantile(effect, .95) ~ 'Top',
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(Range)) %>%
  group_by(Range) %>%
  select(Range, p_irat, black, hse_inc, ccred, mcred, pubrec,
       denpmi, selfemp, single, hischl, ltv_med, ltv_high) %>%
  summarize(across(.fns = mean)) %>%
  pivot_longer(cols = 2:13) %>%
  pivot_wider(id_cols = name, values_from = value, names_from = Range) %>%
  mutate(Ratio = Top/Bottom) %>%
  arrange(Ratio) %>%
  knitr::kable()
```

## Treatment Effect Methods

- Anyway, there's some stuff for you to check out!
- Obviously there are zillions of causal-inference methods we don't have time to cover
- Bartik instruments, matrix completion, causal discovery, and so on and so on and so on
- Consider this a good starting place

## Exam Review

- Just a reminder of some stuff we've covered

## Fixed Effects

- If we have data where we observe the same people over and over, we can implement *fixed effects* by controlling for *individual*
- This accounts for everything that's constant within individual. If, for example, "individual" was city, that would include geography, state, founding year, etc.
- Doesn't account for things that vary within individual over time, like `Laws`

## Difference-in-Difference

- Difference-in-Difference applies when you have a group that you can observe both before and after the policy
- You worry that `time` is a confounder, but you can't control for it
- Unless you add a control group that DIDN'T get the policy
- We must be careful to check that parallel trends holds

## Difference-in-Difference

```{r, dev='CairoPNG', echo=FALSE, fig.width=7, fig.height=5.5}
dag <- dagify(D~Group+Time,
              Y~Group+Time+D,
              coords=list(
                x=c(D=0,Group=1,Time=1,Y=2),
                y=c(D=1,Group=0,Time=2,Y=1)
              )) %>% tidy_dagitty()
ggdag_classic(dag,node_size=20) + 
  theme_dag_blank()
```

## Difference-in-Difference

- Get the before-after difference for both groups
- Then subtract out the difference for the control

```{r, echo=TRUE}
diddata <- tibble(Group=c(rep("C",2500),rep("T",2500)),
                  Time=rep(c(rep("Before",1250),rep("After",1250)),2)) %>%
  mutate(Treated = (Group == "T") & Time == "After") %>%
  mutate(Y = 2*(Group == "T") + 1.5*(Time == "After") + 3*Treated + rnorm(5000))
did <- diddata %>% group_by(Group,Time) %>% summarize(Y = mean(Y))
before.after.control <- did$Y[1] - did$Y[2]
before.after.treated <- did$Y[3] - did$Y[4]
did.effect <- before.after.treated - before.after.control
did.effect
m <- feols(Y ~ Treated | Group + Time, data = diddata)
coef(m)
```

## Regression Discontinuity

- If we have a treatment `D` that is assigned based on a cutoff in a running variable, we can use regression discontinuity
- Focus right around the cutoff and compare above-cutoff to below-cutoff
- We've isolated a great set of treatment/control groups because in this area it's basically random whether you're above or below the cutoff

## Regression Discontinuity

```{r, dev='CairoPNG', echo=FALSE, fig.width=7, fig.height=5.5}
dag <- dagify(D~Above+W,
              Above~Run,
              Run~W,
              Y~D+Run+W,
              coords=list(
                x=c(D=1,Above=2,W=3,Run=3,Y=4),
                y=c(D=1,Above=1.25,W=2,Run=1.25,Y=.5)
              )) %>% tidy_dagitty()
ggdag_classic(dag,node_size=20) + 
  theme_dag_blank()
```

## Regression Discontinuity

- Estimate by fitting a line that jumps at the cutoff and estimating the jump
- Use local regression and bandwidths to avoid being affected by far-away observations
- "Fuzzy" designs where treatment only jumps partially scale the effect using IV

## Regression Discontinuity

- Expressed well in graphs! Treatment should jump at cutoff. If not perfectly from 0% to 100%, use IV too

```{r, dev='CairoPNG', echo=FALSE, fig.width=7, fig.height=5}
rdddata <- tibble(W=rnorm(10000)) %>%
  mutate(run = runif(10000)+.03*W) %>%
  mutate(treated = run >= .6) %>%
  mutate(Y = 2+.01*run+.5*treated+W+rnorm(10000))
ggplot(rdddata,aes(x=run,y=Y,color=treated)) + geom_point()+
  geom_vline(aes(xintercept=.6))+
  geom_smooth(aes(group = treated), size = 2, color = 'black') +
  labs(x='Running Variable',
       y='Outcome') + 
  theme_pubr() + 
  guides(color = FALSE)
```

## Regression Discontinuity

- Variables other than `Y` and treatment shouldn't jump at cutoff - they should be balanced

```{r, dev='CairoPNG', echo=FALSE, fig.width=7, fig.height=5}
ggplot(rdddata,aes(x=run,y=W,color=treated)) + geom_point()+
  geom_vline(aes(xintercept=.6))+
  geom_smooth(aes(group = treated), size = 2, color = 'black') +
  labs(x='Running Variable',
       y='W') +
   theme_pubr() + 
  guides(color = FALSE)
```

## Instrumental Variables

- An instrumental variable affects treatment (relevant) but has no back doors itself or paths to $Y$ except through $X$ (valid) 
- We move the no-open-back-doors assumption to the IV rather than the treatment
- We isolate JUST the variation that comes from `Z`. No back doors in that variation! We have a causal effect
- Can conceptually think of it as (or literally apply it to) an experiment where randomization doesn't work perfectly

## Instrumental Variables

```{r, dev='CairoPNG', echo=FALSE, fig.width=7, fig.height=5.5}
dag <- dagify(X~Z+W,
              Y~X+W,
              coords=list(
                x=c(Z=0,X=1,W=1.5,Y=2),
                y=c(Z=0,X=0,W=1,Y=0)
              )) %>% tidy_dagitty()
ggdag_classic(dag,node_size=20) +
  theme_dag_blank()
```

## Treatment Effects

- There isn't *a* treatment effect. They vary across time, space, individual
- Our methods give us averages - ATE (experiment), ATT (DID), LATE (IV, RDD), variance-weighted (regression w/ controls), etc.
- We must pay close attention to what our design *and estimator* gives us

## That's it!

- In a very condensed way, that's the material we covered!
- I recommend looking back over slides, notes, homeworks
